A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-05-12 paper/paper.Rmd
code not codes

For statistics, and data science by association, it is increasingly important to reach the individual. One might think of this as a divergence of purpose -- statistics for public policy -- or statistics for the public. The two are not the same. As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual. For example, with the wages data, the overall trend, across demographics, is a steady increase in wages, but the patterns among individuals is heterogeneous. Some individuals see a decline in their wages, some an increase, some have dramatic ups and downs from year to year, and the heterogeneity of the patterns are evident across demographics. Subject-specific variation explains most of the variation in wages, and even though the overall trend is statistically significant, it is weak. The message for public policy is that demographic profile is related to different wage patterns on average and some structural social change is desirable. However, discussing the overall trend with an individual is misleading -- more than likely their pattern is completely different. It could even be depressing for them to compare themselves to the overall average, especially if the individual's wages have a declining trend or high volatility. It would be more useful for an individual to know what percentage of the population have profiles like theirs. We argue that different summaries are needed to communicate with the public, and ones where they might be able to individually act on to change their own situation. <!--this information is the appropriate message for an individual is that you as an individual are far more than your demographic. The majority of people in the study do not have a pattern that is similar to the overall trend. If you have experienced declining wages, you are not alone; there are others like you, and more than you think.--> We occasionally hear this voiced in the public media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage]. Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.

@@

\begin{figure*}[p]

\begin{tcolorbox}[title = Navigating the data source, fontupper=\linespread{.8}\selectfont]
\vspace{3mm}
\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\\\
\vspace{3mm}
\faCheck\ The CASEID will be always be selected.  \\
\vspace{1mm}
\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  For the remaining variables, we went to the ``Variable Search" tab and select variables as follows
\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year
\begin{itemize}
\item[$\triangleright$] Highest grade completed
\begin{itemize}
\item[\faCheck] All 80 variables in Highest grade completed were selected.
\end{itemize}
\end{itemize}
\begin{itemize}
\item[$\triangleright$] Dates of diploma or degree
\begin{itemize}
\item[\faCheck] All variables named Q3-8A were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job
\begin{itemize}
\item[$\triangleright$] Hours worked  
\begin{itemize}
\item[\faCheck] All 447 primary variables in Hours worked were selected.
\end{itemize}
\item[$\triangleright$] Hourly wages
\begin{itemize}
\item[\faCheck] All 156 variables in Hourly wages were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ Since date of last interview $\triangleright$ Weeks worked
\begin{itemize}
\item[\faCheck] All 28 variables in Weeks worked were selected.
\end{itemize}
\end{itemize}
\begin{itemize}
\item[$\triangleright$] Employer Roster $\triangleright$ Job dates $\triangleright$ Original start date
\begin{itemize}
\item[\faCheck] Only selected the start date (Year) for the first job (E00101.02)
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography \& Demographics
\begin{itemize}
\item[$\triangleright$] Demographics $\triangleright$ Basic demographics $\triangleright$ Date of birth
\begin{itemize}
\item[\faCheck] All 4 variables in Date of birth were selected. 
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[\faCloudDownload] To download all 742 variables selected, we then navigate to the tab ``Save / Download" then select the tab ``Advanced Download". We select the R Source code and comma-delimited datafile of selected variables with Reference Number as column headers. We name the filename ``NLSY79" and press the download button. There are also options to get control or dictionary files for SAS, SPSS and STATA. 
\end{itemize}
\end{tcolorbox}
\caption{Documented steps taken to select variables of interest and download the raw data.\label{fig:source-nav}}
\end{figure*}

@@

- `NLSY79.csv`: comma separated value format of the response data,
- `NLSY79.dat`: alternative text format of the response data,
- `NLSY79.NLSY79`: tagset of variables that can be uploaded to the website to recreate the data set, and
- `NLSY79.R`: R script for reading the data into R and converting the variable names and label into something more sensible.

@@

<!--AE comment: first sentence can be removed or rephrased. Next sentence includes the word wages too many times. Next sentence should say “dropouts” rather than “drop outs” to be consistent with the previous sentence. -->
The original set, containing wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994, is available in the R package `brolgar`. To compare the refreshed data with the original, we need to subset the data into a matching cohort of high school dropouts. There are multiple ways to match a comparative cohort from the updated data -- the simplest approach is to subset to individuals with matching IDs in the original data and restricting the longitudinal measurements to the same years; another approach is to replicate the process as suggested by the description of the original data thus identifying a new (and possibly removal of old) set of individuals that is included in the refreshed data. We use the latter approach and as the first step we identify individuals who dropped out of high school.

@@

1. An individual whose highest grade completed (`hgc`) is reported to be less than 12th grade, **or**
2. An individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and has received a GED (i.e. `ged` code is 2).

An additional criterion from @SingerJudithD2003Alda is to only include males aged between 14 and 17 years old in 1979. With this filtering, we obtained `r ori_do_nrow` individuals in the refreshed data compared to `r sw_do_nrow` individuals in the original data. To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id. This revealed several reasons for the disparity:

1. `r nrow(join_more_17)` individuals were more than 17 years old in 1979. Thus, it looks like the description of the original data is not quite accurate that there are people older than 17 in the subset. Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later. 
2. `r nrow(join_less_17)` individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
    i. `r nrow(join_not_elig)` of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded as 1). This suggests they are not dropouts, and so we excluded them from the refreshed data. 
    ii. The information about `ged` is missing in `r nrow(join_missing_ged)` individuals. We decided to include them in the refreshed data. 
    iii. `r nrow(join_ged_both)` individuals have both diploma and GED (`ged` is coded as 3). These were kept in the refreshed data. 
    iv. `r nrow(join_not_exist)` individuals do not exist in `wages` data because they have participated in less than 3 rounds of the survey.

The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing `r n_do` individuals.

@@

Figure \ref{fig:compare-subsets} contains summaries of corresponding variables in the two subsets. Plot (A) shows a back-to-back bar chart of the highest grade completed. The two sets are almost the same, but small differences remain. Plots (B) and (C) show stacked density plots of experience and log wages, respectively. The distributions are relatively close, with more differences in wages as would be expected because the refreshed data is not inflation-adjusted.

@@

Ultimately, the refreshed data is reasonably similar to the original but unsatisfactorily far from it. The last step required would be to inflation-adjust wages. This is better to do with each wave of new data added so that it is relative to the last date in the data. Our decision was to provide the raw wages and include code to make the adjustment as part of the package.

@@

Having trustworthy data is imperative for statistics and data science education. Whenever one uses a textbook data set that is perceived as relating to the students' lives, there will be interpretations made. The wages data is an example of this. Students can take away multiple interpretations from the data (e.g. wages increase with experience, education, and race) based on the teaching focus to learn about the world. If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., sex and race), or data that has been poorly processed containing errors, we, as educators, are being irresponsible because the societal message taught to students may be flawed. This paper demonstrates the process of producing a trustworthy data set for teaching.

@@

- **Codes**: R script to reproduce data tidying and cleaning is available at \url{`r yowie_pkgdown`/articles/process-data.html}. The code for the extreme value handling with robust linear fits is in \url{`r yowie_pkgdown`/articles/input-to-valid-data.html}.

>>>>

For statistics, and data science by association, it is increasingly important to reach the individual. One might think of this as a divergence of purpose -- statistics for public policy -- or statistics for the public. The two are not the same. As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual. For example, with the wages data, the overall trend, across demographics, is a steady increase in wages, but the patterns among individuals is heterogeneous. Some individuals see a decline in their wages, some an increase, some have dramatic ups and downs from year to year, and the heterogeneity of the patterns are evident across demographics. Subject-specific variation explains most of the variation in wages, and even though the overall trend is statistically significant, it is weak. The message for public policy is that demographic profile is related to different wage patterns on average and some structural social change is desirable. However, discussing the overall trend with an individual is misleading -- more than likely their pattern is completely different. It could even be depressing for them to compare themselves to the overall average, especially if the individual's wages have a declining trend or high volatility. It would be more useful for an individual to know what percentage of the population have profiles like theirs. We argue that different summaries are needed to communicate with the public, and ones where they might be able to individually act on to change their own situation. <!--this information is the appropriate message for an individual is that you as an individual are far more than your demographic. The majority of people in the study do not have a pattern that is similar to the overall trend. If you have experienced declining wages, you are not alone; there are others like you, and more than you think.--> We occasionally hear this voiced in the public media media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage]. Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.

@@

\begin{figure}[t]

\begin{tcolorbox}[title = Navigating the data source]
\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
\vspace{1mm}
\faCheck\ The CASEID will be always be selected.  \\
\vspace{1mm}
\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  \\
\vspace{1mm}
For the remaining variables, we went to the "Variable Search" tab and select variables as follows
\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year
\begin{itemize}
\item[$\triangleright$] Highest grade completed
\begin{itemize}
\item[\faCheck] All 80 variables in Highest grade completed were selected.
\end{itemize}
\end{itemize}
\begin{itemize}
\item[$\triangleright$] Dates of diploma or degree
\begin{itemize}
\item[\faCheck] All variables named Q3-8A were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job
\begin{itemize}
\item[$\triangleright$] Hours worked  
\begin{itemize}
\item[\faCheck] All 447 primary variables in Hours worked were selected.
\end{itemize}
\item[$\triangleright$] Hourly wages
\begin{itemize}
\item[\faCheck] All 156 variables in Hourly wages were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ Since date of last interview $\triangleright$ Weeks worked
\begin{itemize}
\item[\faCheck] All 28 variables in Weeks worked were selected.
\end{itemize}
\end{itemize}
\begin{itemize}
\item[$\triangleright$] Employer Roster $\triangleright$ Job dates $\triangleright$ Original start date
\begin{itemize}
\item[\faCheck] Only selected the start date (Year) for the first job (E00101.02)
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography \& Demographics
\begin{itemize}
\item[$\triangleright$] Demographics $\triangleright$ Basic demographics $\triangleright$ Date of birth
\begin{itemize}
\item[\faCheck] All 4 variables in Date of birth were selected. 
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[\faCloudDownload] To download all 742 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download". We select the R Source code and comma-delimited datafile of selected variables with Reference Number as column headers. We name the filename "NLSY79" and press the download button. There are also options to get control or dictionary files for SAS, SPSS and STATA. 
\end{itemize}
\end{tcolorbox}
\caption{Documented steps taken to select variables of interest and download the raw data.\label{fig:source-nav}}
\end{figure}

@@

- `NLSY79.csv`: comma separated value format of the response data,
- `NLSY79.dat`: alternative text format of the response data,
- `NLSY79.NLSY79`: tagset of variables that can be uploaded to the website to recreate the data set, and
- `NLSY79.R`: R script for reading the data into R and converting the variables' names and label into something more sensible.

@@

<!--AE comment: first sentence can be removed or rephrased. Next sentence includes the word wages too many times. Next sentence should say “dropouts” rather than “drop outs” to be consistent with the previous sentence. -->
The original set, containing wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994, is available in the R package `brolgar`. To compare the refreshed data with the original, a subset needs to be matched. There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data and restricting the longitudinal measurements to the same years. However, we decided to try to replicate the process, as suggested by the description of the original data. This requires first identifying individuals who dropped out of high school.

@@

1. An individual whose highest grade completed (`hgc`) is reported to be less than 12th grade, **or**
2. An individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and has received a GED (`ged` is code to 2).

An additional criterion from @SingerJudithD2003Alda is to only include males aged between 14 and 17 years old in 1979. With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data. To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id. This revealed several reasons for the disparity:

1. 173 individuals were more than 17 years old in 1979. Thus, it looks like the description of the original data is not quite accurate that there are people older than 17 in the subset. Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later. 
2. 79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
    i. 35 of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded to 1). This suggests they are not dropouts, and so we excluded them from the refreshed data. 
    ii. The information about `ged` is missing in 38 individuals. We decided to include them in the refreshed data. 
    iii. 3 individuals have both diploma and GED (`ged` is coded to 3). These were kept in the refreshed data. 
    iv. 12 individuals do not exist in `wages` data because they have participated in less than 3 rounds of the survey.

The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing 863 individuals.

@@

Figure \ref{fig:compare-subsets} contains summaries of corresponding variables in the two subsets. Plot (A) shows a back-back bar chart of the highest grade completed. The two sets are almost the same, but small differences remain. Plots (B) and (C) show stacked density plots of experience and log wages, respectively. The distributions are relatively close, with more differences in wages as would be expected because the refreshed data is not inflation-adjusted.

@@

Ultimately, the refreshed data is reasonably similar to the original but unsatisfactorily far from it. The last step required would be to inflation-adjust wages. This is better to do with each wave of new data added so that it is relative to the last date in the data. Our decision was to provide the raw wages and include codes to make the adjustment as part of the package.

@@

Having trustworthy data is imperative for statistics and data science education. Whenever one uses a textbook data set that is perceived as relating to the students' lives, there will be interpretations made. The wages data is an example of this. Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race. If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., sex and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed. This paper demonstrates the process of producing a trustworthy data set for teaching.

@@

- **Code**: R script to reproduce data tidying and cleaning is available at \url{`r yowie_pkgdown`/articles/process-data.html}. The code for the extreme value handling with robust linear fits is in \url{`r yowie_pkgdown`/articles/input-to-valid-data.html}.
